Indexing Variation Graphs\footnote{This work was supported by the Wellcome Trust grant [098051].}

### 2016-04-19 paper/paper.tex
Almost finished the paper. Probably full of typos and confusion.

1000GP: \cite{1000GP2015}

Graph genomes: \cite{Schneeberger2009}

FM-index: \cite{Ferragina2005a}

XBW: \cite{Ferragina2009b}

GCSA: \cite{Siren2014}

Succinct DBG: \cite{Bowe2012}, variable-order \cite{Boucher2014}, alternative representation \cite{Roedland2013}

BWBBLE: \cite{Huang2013}

Contracted DBG: \cite{Cazaux2014}

Compressed DBG: \cite{Marcus2014}

Hypertext index: \cite{Thachuk2013}

FM-index of alignment: \cite{Na2015}

HISAT / HISAT2: \cite{Kim2015}

Mention that proofs of lemmas are in the appendix

@@

\section{Path indexes}\label{sect:path-indexes}

@@

@@

De~Bruijn graphs often have nodes that are redundant when the graph is used as a path index. Removing these nodes from the graph would make the index smaller and more practical. On the other hand, we still want to guarantee that the index works correctly with patterns of length $k$ or less, where $k$ is the order of the de~Bruijn graph.

@@

Pruned de~Bruijn graphs are similar to manifold de Bruijn graphs \cite{Lin2014}.

@@

\begin{lemma}[Pruning]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$. If $\abs{V'_{K}} > 0$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph remains an \orderk{k} pruned de~Bruijn graph of $G$, if we remove $G'.\gkey(v)$ for all $v \in V'_{K}$ from the key set and replace them with $K$.
\end{lemma}

We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold. Let $G_{d} = (V_{d}, E_{d})$ be the \orderk{k} de~Bruijn graph of graph $G$, and let $G' = (V', E')$ be a pruned de~Bruijn graph produced by that process. Each node $v' \in V'$ is an \emph{equivalence class} of nodes $V_{d}(v') \subseteq V_{d}$ corresponding to a shared prefix $G'.\gkey(v')$ of keys. For all $v \in V_{d}(v')$, we have $G_{d}.\gvalue(v) = G'.\gvalue(v')$.

@@

\begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned \orderk{k} de~Bruijn graph. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all nodes $v' \in V'$, characters $c \in \Sigma$, and nodes $u' \in G'.\gpred(v', c)$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

@@

Path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}. As our technical details are different from the ones used in that paper, we now describe how the GCSA of a path graph works.

@@

\subsection{Index construction}

@@

% FIXME justify why we can use parent queries
% we need a lemma stating that we always match entire equivalence classes
If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. However, if pattern $X$ of length $\abs{X} \le k$ matches lexicographic range $[sp, ep]$, we can safely use a $\parent$ query for the corresponding node in the suffix tree.

\section{Experiments}

The \emph{variation} graphs used in vg have strings as node labels. The average length of a node label is typically around 10~bases. Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to first offsets of variation graph nodes are always selected as samples for performance reasons.

A single-strand whole-genome index can be built overnight with 96~gigabytes of memory. Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower. The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower. The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.

We extracted \kmer{16}s and \kmer{32}s from the vg graphs for benchmarking purposes using \texttt{vg sim}. We then queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. The main reason is that we use to use $\select$ queries, which are much slower than $\rank$ queries on plain bitvectors \cite{Gog2014}.

\begin{table}[t!]
\begin{center}
\caption{Performance benchmarks using an \orderk{64} GCSA2. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct values; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in milliseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Matches} & \textbf{Nodes} & \textbf{Occs} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1830M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 0.47 ms \\
\hline
\end{tabular}
\end{center}
\end{table}

\section{Discussion}

\clearpage
\appendix
\section{Proofs of lemmas}

\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
(a) The key $G.\gkey(u')$ of a node $u' \in G'.\gpred(v', c)$ prefix-matches string $c \cdot G.\gkey(v')$. If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G.\gkey(v)$ in order to be distinct. In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G.\gkey(v')$ and $G.\gkey(w')$ would prefix-match. Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node. Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.

(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$. Key $G'.\gkey(u')$ must have $c \cdot G'.\gkey(v')$ as a prefix. If there is a substring $S[i,j] = c \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', c)$. By (a), $u'$ is the only such node, and hence string $c \cdot G'.\gkey(v')$ can also serve as its key.

(c) If $\abs{X} = 0$, every key prefix-matches the pattern. If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$. Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$. By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$. By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$. Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.

Note that any key set satisfying (a) and (b) also satisfies (c). If $G'' = (V'', E'')$ is a pruned de~Bruijn graph and $v'' \in V''$ is a node, any pattern matching node $v''$ prefix-matches the key of the corresponding equivalence class in graph $G'$.
\end{proof}

>>>>

Sequencing pipelines typically start with mapping the reads from the sequenced genome to a \emph{reference genome} of the same species. As reference genomes are usually assembled from the genomes of a small number of individuals, they are biased towards those individuals. This \emph{reference bias} may affect the results of subsequent analysis, especially when the sequenced individuals are from different populations than the ones behind the reference genome.

Recently there has been much interest in \emph{variation graphs} (also called pan-genomes, graph genomes, or reference graphs), which encode the genetic variation within a population as a graph \cite{Paten2014,Marcus2014,Church2015,Dilthey2015,Marschall2016}. These graphs are eventually expected to replace reference sequences as reference genomes. The shift to graph references will likely take years, as new methods and tools must be developed to replace those based on linear references.

The \emph{variation graph toolkit vg} \cite{Garrison2014-2016} is a community effort to develop such tools. This paper describes \emph{GCSA2}, the path index developed for vg. A \emph{path index} is a generalization of text indexes for labeled graphs. They are used for finding paths with labels that match the query string. Indexing graphs for path queries is an inherently hard problem, as the number of paths increases exponentially with path length. Hence the design of a path index is a trade-off between maximum supported query length, the size of the index, and pruning complex regions of the graph before indexing.

Mapping reads to a graph was first investigated by Schneeberger et~al.~\cite{Schneeberger2009}. Sirén et~al.~\cite{Siren2014} developed \emph{GCSA}, which generalized the FM\nobreakdash-index \cite{Ferragina2005a} from a text index to a path index. The original GCSA could only index directed acyclic graphs and depended on pruning the complex regions, as it did not place any limits on query length. Kim et~al.~\cite{Kim2015-2016} combined GCSA with the ideas developed for the HISAT read aligner \cite{Kim2015}. The resulting HISAT2 aligner was the first practical graph-based read aligner, though it only uses the graph for more accurate mapping to a linear reference.

The \emph{succinct de~Bruijn graphs} of Bowe et.~al.~\cite{Bowe2012} used a similar generalization of the FM\nobreakdash-index as in GCSA for representing \emph{de~Bruijn graphs} space-efficiently. Rødland \cite{Roedland2013} proposed another similar generalization. The succinct de~Bruijn graphs have been generalized to represent \orderk{k} de~Bruijn graphs with different values of $k$ in the same structure \cite{Boucher2014}, but they still need to store the graph for the largest value of $k$ explicitly. Compacted \cite{Cazaux2014} and compressed de~Bruijn graphs \cite{Marcus2014} achieve similar space savings by representing unary paths in the graph as a single node. A prototype read aligner based on compacted de~Bruijn graphs has been proposed \cite{Limasset2015}.

Other approaches to indexing variation graphs include BWBBLE \cite{Huang2013}, which indexes the graph as a collection of sequences. Simple variants are encoded by using the powerset alphabet, while more complex variants are represented as separate sequences, with a sufficient amount of context around the variant site. This is usually enough to keep the index small, at the price of much higher query times. The hypertext index \cite{Thachuk2013} is another similar idea. The nodes of the graph are labeled with strings, which are indexed using an ordinary FM-index. Partial matches in the strings are combined into full matches with range queries in the edge matrix. While matches crossing one edge are easy to find, the approach becomes complex and potentially slow with matches crossing multiple edges. There are also several indexes (e.g.~\cite{Huang2010,Wandelt2013,Danek2014,Na2015}), which use a graph as a space-efficient representation for similar sequences but index the sequences instead of the graph.

GCSA2 combines ideas from the original GCSA and from succinct de~Bruijn graphs. Conceptually it uses a de~Bruijn graph as a path index of a variation graph. The \orderk{k} de~Bruijn graph is \emph{pruned} (compressed structurally) by using strings shorter than $k$ characters as nodes, if the shorter strings identify the start nodes of the corresponding paths uniquely. The pruned graph is then encoded with a generalization of the FM\nobreakdash-index. GCSA2 includes novel extensions based on what is essentially a suffix tree of the pruned de~Bruijn graph. These extensions are used for e.g.~finding maximal exact matches in the vg read aligner.

This paper contains several lemmas describing the properties of generalizations of de~Bruijn graphs. Their proofs are included in Appendix~\ref{appendix:proofs}.

@@

\section{Path Indexes}\label{sect:path-indexes}

@@

\begin{lemma}[Context length]\label{lemma:pg-context}
Let $G' = (V', E')$ be a path graph, and let $X \in \patternset$ be a pattern. The set $G'.find(X)$ consists of all nodes $v' \in V'$ such that $X[0, m-1]$ is a prefix of $G'.\gkey(v')$, for a context length $m$.
\end{lemma}

@@

De~Bruijn graphs often have nodes that are redundant when the graph is used as a path index. By carefully pruning the graphs, we get smaller graphs that are equivalent to \orderk{k} de~Bruijn graphs with patterns of length $k$ or less. These pruned de~Bruijn graphs naturally arise from GCSA construction. They are similar to manifold de Bruijn graphs \cite{Lin2014}.

@@

@@

\begin{lemma}[Pruning]\label{lemma:dbg-prune}
Let $G = (V, G)$ be a graph, let $G' = (V', E')$ be the \orderk{k} pruned de~Bruijn graph of $G$ with key set $\mathcal{K}$, let $K \in \Sigma^{\ast}$ be a string of length $\abs{K} > 0$, and let $V'_{K}$ be the set of nodes $v \in V'$ having string $K$ as a prefix of $G'.\gkey(v)$.
If $\abs{V'_{K}} > 1$ and $G'.\gvalue(u') = G'.\gvalue(v')$ for all $u, v \in V'_{K}$, the path graph with key set $(\mathcal{K} \cup \set{K}) \setminus V'_{K}$ is an \orderk{k} pruned de~Bruijn graph of $G$.
\end{lemma}

We can compress a de~Bruijn graph structurally by repeatedly merging sets of nodes sharing a common prefix of their keys, as long as the conditions of Lemma~\ref{lemma:dbg-prune} hold. Let $G' = (V', E')$ be an \orderk{k} pruned de~Bruijn graph, and let $G'' = (V'', E'')$ be the same graph after further pruning. Each node $v'' \in V''$ is an \emph{equivalence class} of nodes $V'(v'') \subseteq V'$ corresponding to a shared prefix $G''.\gkey(v'')$ of keys. For all $v' \in V'(v'')$, we have $G'.\gvalue(v') = G''.\gvalue(v'')$.

@@

\begin{lemma}[Maximal pruning]\label{lemma:dbg-maximal}
Let $G' = (V', E')$ be a maximally pruned de~Bruijn graph of graph $G = (V, E)$. Then
(a) $\abs{G'.\gpred(v', c)} \le 1$ for all nodes $v' \in V'$ and characters $c \in \Sigma$;
(b) $\abs{G'.\gkey(u')} \le \abs{G'.\gkey(v')}+1$ for all edges $(u', v') \in E'$; and
(c) key $G'.\gkey(v')$ prefix-matches pattern $X \in \patternset$ for all nodes $v' \in G'.\find(X)$.
\end{lemma}

@@

The path graphs are essentially equivalent to the prefix-sorted automata, which were used in the original paper on GCSA \cite{Siren2014}. As many of the technical details are different, we now take a look at how a GCSA of a path graph works. We also extend GCSA with techniques based on suffix trees. Appendix~\ref{appendix:construction} describes the construction algorithm used with GCSA2. While the algorithm is a similar prefix-doubling algorithm as in the original GCSA, it keeps the paths and graphs on disk to avoid excessive memory usage.

@@

@@

If we build an $x$\nobreakdash-ary tree over the LCP array of graph $G'$, with each internal node storing the minimum LCP value in the corresponding range of the LCP array, we can support the required queries in $\Oh{x + \log_{x} \abs{V'}}$ time using $\frac{x}{x-1} \abs{V'} \log k$ bits of space. Because nodes of a path graph may match patterns that do not prefix-match their keys, we have to be careful with the suffix tree operations we use. By Lemma~\ref{lemma:pg-context}, the ranges returned by $\find$ queries always correspond to prefixes of the pattern. Hence we can safely use $\parent$ queries with such ranges.

\section{Implementation and Experiments}

The \emph{variation graphs} used in vg have strings as node labels. The average length of a node label is typically around 10~bases. Each node can be traversed both in forward orientation and in reverse complement orientation, and edges may cross from the forward strand of one node to the reverse complement strand of another node, or the other way around. When extracting paths from a variation graph, the graph is implicitly converted into an input graph with single-character labels. Input graph nodes corresponding to the initial offsets of variation graph node labels are always selected as samples for performance reasons.

A single-strand whole-genome index can be built overnight with 96~gigabytes of memory. Even 64~gigabytes is enough, but the lack of sufficient disk cache may make the construction slower. The size of the de~Bruijn graph grows exponentially at roughly $1.031^{k} \cdot 2.38$~billion nodes, while the growth rate of the pruned de~Bruijn graph is much slower.\footnote{The merging step, as currently implemented, does not produce a maximally pruned graph. It consists of a pruning step, followed by merging the paths with identical labels.} The index itself grows even more slowly with $k$, as we sample the values more often in lower order graphs.

We extracted \kmer{16}s, \kmer{32}s, and \kmer{64}s from the vg graphs using \texttt{vg sim}, and queried for them in the \orderk{64} index using a single thread. The results can be seen in Table~\ref{table:benchmark}. Not all patterns could be matched to the graphs they were extracted from, as some paths were pruned during construction. The $\find$ and $\locate$ queries are 5\nobreakdash--10 times slower than in a fast whole-genome FM\nobreakdash-index. There are two main reasons: (1) we use $\select$ queries, which are slower than $\rank$ queries on plain bitvectors \cite{Gog2014}; and (2) $\locate$ queries generate duplicates, which must be filtered out. Switching to the simplified encoding for maximally pruned graphs would make $\find$ queries faster and $\locate$ queries slower.

\begin{table}[t!]
\begin{center}
\caption{Performance benchmarks using an \orderk{64} GCSA2. Pattern length; number of patterns; number of matching patterns; number of matching nodes; number of distinct occurrences; average time for $[sp, ep] = \find(X)$, $\parent(sp, ep)$, and $\countq(sp, ep)$ queries in microseconds; and average time for $\locate(sp, ep)$ queries in microseconds or milliseconds.}\label{table:benchmark}
\begin{tabular}{c|cc|cc|cccc}
\hline
\textbf{Length} & \textbf{Patterns} & \textbf{Occs} & \textbf{Nodes} & \textbf{Values} & $\find()$ & $\parent()$ & $\countq()$ & $\locate()$ \\
\hline
16 & 351584 & 347453 & 1829M &  872M & 18.1 \textmu{}s & 0.40 \textmu{}s & 0.94 \textmu{}s & 26.2 ms \\
32 & 351555 & 333258 & 52.6M & 34.3M & 37.9 \textmu{}s & 0.28 \textmu{}s & 0.43 \textmu{}s & 468 \textmu{}s \\
64 & 351567 & 326101 & 0.33M & 1.35M &  142 \textmu{}s & 0.34 \textmu{}s & 0.57 \textmu{}s & 12.4 \textmu{}s \\
\hline
\end{tabular}
\end{center}
\end{table}

\section{Conclusions}

GCSA2 is a path index for variation graphs. It uses a structurally compressed de~Bruijn graph of the paths in the input graph as an index, and encodes that with a generalization of the FM\nobreakdash-index. The index also incorporates extensions based on a generalization of suffix trees. GCSA2 has been integrated in the variation graph toolkit vg, where it is used e.g.~for read alignment based on maximal exact matches.

We can build a forward strand whole-genome index overnight on a system with 96~gigabytes of memory and a few hundred gigabytes of fast disk space. The index itself requires less than 16~gigabytes of memory. Query performance is roughly 5\nobreakdash--10 times slower than for a fast FM\nobreakdash-index for sequences. GCSA2 can search for patterns at roughly 0.8\nobreakdash--1.0~MB/s and locate 100,000\nobreakdash--300,000 occurrences per second using a single CPU core. As with other FM\nobreakdash-indexes, the speedup from distributing the queries on multiple cores is quite high.

The compressed suffix tree used in GCSA2 may be useful for indexing texts as well, as it is both simple to implement and fast in practice. As long as the average LCP value is low, we can generalize the byte array encoding of the LCP array \cite{Abouelhoda2004} to implement it space-efficiently. We use an array of $m$\nobreakdash-bit integers for the LCP array and the $x$\nobreakdash-ary tree, mark the values that are too large for $m$~bits with the largest possible value, and store them separately.

% FIXME from here
% additional edges using hypertext index; haplotypes and potential recombinations

% graph formalism, grammars

\clearpage
\appendix
\section{Proofs of Lemmas}\label{appendix:proofs}

\begin{proof}[Proof of Lemma~\ref{lemma:pg-context} (Context length)]
If $\abs{X} \le 1$, the statement is true by definition for $m = \abs{X}$. Now assume that $M_{i+1} = G'.\find(X[i+1, \abs{X}-1])$ is the set of all nodes $v' \in V'$ such that substring $X[i+1, i+m]$ is a prefix of $G'.\gkey(v')$.

Consider the set $M_{i} = \bigcup_{v' \in M_{i+1}} G'.\gpred(v', X[i])$, and assume that it is nonempty. Edge $(u', v') \in E'$ exists if and only if key $G'.\gkey(u')$ prefix-matches string $G'.\glabel(u') \cdot G'.\gkey(v')$. Hence $u' \in M_{i}$ if and only if key $G'.\gkey(u')$ prefix-matches string $X[i, i+m]$. If $\abs{M_{i}} > 1$, string $X[i, i+m]$ is a proper prefix of key $G'.\gkey(u')$ for all nodes $u' \in M_{i}$ due to the prefix-free property, and we can continue with context length $m+1$. Otherwise the context length becomes $\min(m+1, \abs{G'.\gkey(u')})$ for the only node $u' \in M_{i}$.
\end{proof}

\begin{proof}[Proof of Lemma~\ref{lemma:dbg-maximal} (Maximal pruning)]
(a) For each node $u' \in G'.\gpred(v', c)$, the key $G'.\gkey(u')$ prefix-matches string $c \cdot G.\gkey(v')$. If $\abs{G'.\gpred(v', c)} > 1$, the keys must be longer than string $c \cdot G'.\gkey(v')$ or they will prefix-match one another. In that case, there cannot be an edge $(u', w')$ to another node $w' \ne v'$, as the keys $G'.\gkey(v')$ and $G'.\gkey(w')$ would prefix-match. Hence $G'.\gvalue(u')$ must be the union of sets $G.\gpred(v, c)$ over nodes $v \in G'.\gvalue(v')$, and we can merge the set $G'.\gpred(v', c)$ into a single node. Because graph $G'$ is already maximally pruned, it must be that $\abs{G'.\gpred(v', c)} \le 1$.

(b) Assume that $\abs{G'.\gkey(u')} > \abs{G'.\gkey(v')}+1$. Then key $G'.\gkey(u')$ must have string $G'.\glabel(u') \cdot G'.\gkey(v')$ as a prefix. If there is a substring $S[i,j] = G'.\glabel(u') \cdot G'.\gkey(v')$ in the \kcollection{k} used for building graph $G'$, the key starting at $S[i]$ must belong to a node in $G'.\gpred(v', G'.\glabel(u'))$. By (a), $u'$ is the only such node, and there cannot be an edge $(u', w') \in E'$ to a node $w' \ne v'$. Hence string $G'.\glabel(u') \cdot G'.\gkey(v')$ can also serve as the key of node $u' \in V'$.

(c) If $\abs{X} = 0$, every key prefix-matches the pattern. If $\abs{X} = 1$, set $G'.\find(X)$ is the set of nodes $v' \in V'$ with $G'.\glabel(v') = X[0]$. Now assume that $\abs{X} \ge 2$ and that $G'.\gkey(v')$ prefix-matches $X[1, \abs{X}-1]$ for all $v' \in G'.\find(X[1, \abs{X}-1])$. By (a), there is at most one node $u' \in G'.\gpred(v', X[0])$ for each such node $v'$. By (b), key $G'.\gkey(u')$ is a prefix of $X[0] \cdot G'.\gkey(v')$. Because $X[1, \abs{X}-1]$ prefix-matches $G'.\gkey(v')$, pattern $X$ prefix-matches string $X[0] \cdot G'.\gkey(v')$, and hence the pattern must prefix-match key $G'.\gkey(u')$.
\end{proof}

\section{Index Construction}\label{appendix:construction}

@@
